Generating Empirically Optimized Numerical Software from MATLAB Prototypes

### 2008-09-19 doc/papers/cgo08/paper.tex
done for tonight
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@351 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
current version
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@353 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\maketitle

@@

\doublespacing
\section{Introduction}
\label{sec:intro}

@@

%Modified from recent DOE proposal

@@

A number of source or binary transformation tools for performance-improving optimizations exist. LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.  A new tool, POET~\cite{POET} also supports a number of loop transformations. POET offers a complex template-based syntax for defining transformations in a language-independent manner. Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler,  telescoping languages~\cite{telescopingurl,teleoverview,Ken99}, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops. PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron. Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations. The transformations target both improved cache locality and parallelism.

@@

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.    To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously. The result is an automatically generated code containing transformed loops and OpenMP directives for compiler parallelization.  Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.
%Does this repeat abstract/intro too closely?

%Probably need some words about WHY this 3 pronged approach is a good idea.

@@

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and MFLOP/s (right) for the VADD operation using different code versions.}
\label{fig:atax}
\end{figure}

@@

The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.

@@

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.3\textwidth}
\footnotesize
\begin{verbatim}
GEMVER
in
  A : column matrix,
  u1 : vector, u2 : vector,
  v1 : vector, v2 : vector,
  a : scalar, b : scalar,
  y : vector, z : vector
out
  B : column matrix,
  x : vector, w : vector
{
  B = A + u1 * v1' + u2 * v2'
  x = b * (B' * y) + z
  w = a * (B * x)
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
\label{fig:gesummv}
\end{figure}

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the BiCG kernel operation using different code versions.}
\label{fig:gesummv}
\end{figure}

@@

fully automate the process to get rid of the manual steps along the way

extend beyond composed BLAS

>>>>

\maketitle
\doublespacing

@@

\section{Introduction}
\label{sec:intro}

@@

Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler, include limited local optimizations for matrix expressions, but do not perform optimizations such as loop fusion across multiple operations as we do with the tools described in this paper. The telescoping languages project~\cite{telescopingurl,teleoverview,Ken99} uses techniques such as strength reduction, vectorization, and procedure specialization to optimize MATLAB scripts, but do not generate reusable optimized linear algebra routines as described in this paper.

@@

A number of source or binary transformation tools for performance-improving optimizations exist. LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.  A new tool, POET~\cite{POET} also supports a number of loop transformations. POET offers a complex template-based syntax for defining transformations in a language-independent manner. Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops. PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron. Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations. The transformations target both improved cache locality and parallelism.

@@

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.  We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously. Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters. Orio then performs an empirical search to select the best among the multiple optimized code variants.

%Probably need some words about WHY this 3 pronged approach is a good idea.
MATLAB optimizing compilers traditionally have not been able to achieve

@@

% Explain how Orio is used

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and MFLOP/s (right) for the VADD operation using different code versions.}
\label{fig:vadd}
\end{figure}

@@

The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:atax}.

@@

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.3\textwidth}
\footnotesize
\begin{verbatim}
GEMVER
in
  A : column matrix,
  u1 : vector, u2 : vector,
  v1 : vector, v2 : vector,
  a : scalar, b : scalar,
  y : vector, z : vector
out
  B : column matrix,
  x : vector, w : vector
{
  B = A + u1 * v1' + u2 * v2'
  x = b * (B' * y) + z
  w = a * (B * x)
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
\label{fig:gemver}
\end{figure}

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the BiCG kernel operation using different code versions.}
\label{fig:bicgkernel}
\end{figure}

@@

The positive initial results from our three-pronged approach to generating tuned linear algebra routines motivate several future lines of investigation. At present there are a number of manual steps involved in obtaining a compiled tuned linear algebra library from the high-level extended MATLAB descriptions. We will next delve into each manual step and create the software that automates it and in effect serves as the ``glue'' between the different tools. We will also investigate extending the supported functionality beyond composed BLAS operations, thus benefiting a wider spectrum of applications.
